---
title: "Lab 2 - Becoming Familiar with R"
output:
  html_document:
    df_print: paged
  word_document: default
---

## Some simple Math


### Use R like a calculator
```{r}
3 + 9 + 12 - 7
12 + 17/2 - 3/4 * 2.5
(12 + 17/2 - 3/4) * 2.5
```
### Try it out - Type some Math

```{r}
pi * 2^3 - sqrt(4)
abs(12-17*2/3-9)
factorial(4)
log(2,10)
log(2,base=10)
log10(2)
log(2)
exp(0.6931472)
log10(2)
10^0.30103
sin(45 * pi / 180)
asin(0.7071068) * 180 / pi
```

### Storing the results of calculations

```{r}
ans1 = 23 + 14/2 - 18 + (7 * pi/2)
ans1
ans2 = 13 + 11 + (17 - 4/7)
ans1 + ans2 / 2
ans3 = ans2 + 9 - 2 + pi
ans4 <- 3 + 5
ans5 <- ans1 * ans2
ans3 + pi / ans4 -> ans6

ans1
ans2
ans3
ans4
ans5
ans6
```


## Reading and Getting Data into R

### Using the combine Command for Making Data

#### Entering Numerical Items as Data
```{r}
data1 = c(3, 5, 7, 5, 3, 2, 6, 8, 5, 6, 9)
data1
```
Here we create a new data set by appending additional values to data 1. It's 
nice that you don't have to expand or concatenate the dataset like in other 
programming languages.  (ex. js - data2 = [...data1,4,5,7,3,4])

```{r}
data2 = c(data1, 4, 5, 7, 3, 4)
data2
```
Here we update the data1 dataset by prepending items and re-assigning to the 
data 1 variable.

```{r}
data1 = c(6, 7, 6, 4, 8, data1)
data1
```
#### Entering Text Items as Data

Below is an example of creating an array of text data.

```{r}
day1 = c('Mon', 'Tue', 'Wed', 'Thu')
day1
```

Mixing text and numbers coerces the numeric data to text.
```{r}
mix = c(data1, day1)
mix
```
### Using the scan Command for Making Data

This does not work great in a notebook - running the below chunk will prompt 
you in R Studio's console. Screenshots will be included in the lab report.
Note that entering a non-numeric throws an error!

```{r}
data3 = scan()
#6 7 8 7 6 3 8 9 10 7
#6 9
data3
```
#### Entering Text as Data

Again, this will prompt you at the command line. Note that entering numeric
data will not throw an error, but instead coerce the data to text.

```{r}
day2 = scan(what='character')
#Mon Tue Wed
#Thu
day2
```

### Using the Clipboard to Make Data

We copy the data from the csv file and paste into the console prompt generated
by this chunk.

```{r}
data4 = scan(sep = ',')
#23,17,12.5,11,17,12,14.5,9
#11,9,12.5,14.5,17,8,21
data4
```
We do the same for text data.

```{r}
data5 = scan(sep = ',', what = 'char')
#"Jan","Feb","Mar","Apr","May","Jun"
#"Jul","Aug","Sep","Oct","Nov","Dec"
data5
```
### Reading a File of Data from a Disk

You can get the present working directory with the below command.

```{r}
getwd()
```

Using the text file I created in this project, I can use a relative path to 
load it in with scan. Note that we have to include a separator to read 
individual values correctly.

```{r}
data6 = scan(sep = ',', file = 'input/cfichman-chap2-numeric-input.txt')
data6
```
You can specify a working directory with the setwd() command. In notebooks, 
the working directory state is contained to each running chunk, which is good, 
because we don't want to mess up the root of our project, but we can still 
demonstrate how setwd() works. This needed to be disabled for HTML report.

```{r}

#setwd('..')
#getwd()

```
You can perform various other filesystem commands with r.

Directory view - list folders AND files, including hidden files with all.files.
```{r}
cwd_contents = dir()
all_cwd_contents = dir(all.files = TRUE)
desktop_contents = dir('Desktop')
CTU_courses = dir('../../')

print("CWD:")
print(cwd_contents)
print("ALL CWD:")
print(all_cwd_contents)
print("Desktop: ")
print(desktop_contents)
print("CTU: ")
print(CTU_courses)

```

List only files
```{r}
list.files()
```
Prompt user to select a file, opening a browser at a the current working 
directory location.
```{r}
data7 = scan(sep=',', file.choose())
data7
```
Scan text data!
```{r}
data8 = scan('input/cfichman-chap2-text-input.txt', what = 'char', sep = ',')
data8
```
## Reading Bigger Data Files

### The read.csv() Command

I've created a file using table2-2 from the text, and made it comma separated.
Now we will read it in.

```{r}
fw = read.csv('input/cfichman-chap2-table-2-2.csv')
fw
```

####Alternative Commands for Reading Data in R

Note that read.table must have header set to true if the input file has a header
and you want to use it. Note it seems like there was a bad edit in the text
and the example doesn't line up with the description, so I followed the example
which shows how to include row names.

```{r}
ssv_path = 'input/cfichman-chap2-space-sep.csv'
ssv = read.table(ssv_path, header = TRUE)
ssv = read.csv(ssv_path, sep = ' ')
ssv
tsv_path = 'input/cfichman-chap2-tsv.csv'
tsv = read.delim(tsv_path)
tsv = read.csv(tsv_path, sep = '\t')
tsv = read.table(tsv_path, header=TRUE, sep = '\t')
tsv

csv2_path = 'input/cfichman-chap2-csv2.csv'
csv2_data = read.csv(csv2_path, row.names = 1, sep = '\t')
csv2_data
```

#### Missing Values in Data Files

```{r}
grass = read.csv('input/cfichman-chap2-na.csv', row.names = 1, sep = '\t')
grass
```
## Viewing Named Objects

### Viewing previously Loaded Named Objects

#### Viewing All Objects
```{r}
ls()
```

#### Viewing Only Matching Names
```{r}
ls(pattern = 'd')
ls(pattern = 'da')
```
Pattern examples. This is similar to Regexp, but lacks much of it's valuable
functionality.

```{r}
ls(pattern = '^d') # only data starting with d
ls(pattern = '1$') # only data ending with a 1
ls(pattern = '^d|^c') #only data starting with a d or a c
ls(pattern = 'data.') # wildcard for one place
```
### Removing Objects from R

You can remove single objects by using their variable names, or you can pass a
list of named objects. Here we're going to clean out all our ^list.$ variables.

```{r}
list = c(1,2,3,4)
list1 = c("hi","hello")
list2 = c(5,6,7)
ls(pattern = 'list')
rm(list)
ls(pattern = '^list.$')
remove(list = ls(pattern = '^list.$'))
ls(pattern = '^list.$')
```
## Types of Data Items

Here we read in a mix of data that interprets which data type to use.

```{r}
grass2 = read.csv('input/cfichman-chap2-grass.csv', sep='\t', as.is = 2)
grass2
```

### Converting between Number and Text data

Here we convert the data we loaded from a file to various types.
```{r}
data7i = as.integer(data7)
data7i
data7n = as.numeric(data7i)
data7n
data7c = as.character(data7)
data7c
data7nt = as.numeric(data7c)
data7nt
```
Now we try with a factor
```{r}
cut_data = c(grass2['cut'])
cut_data
cut_n = as.numeric(as.factor(cut_data))
cut_n
```
Converting from text to numeric results in NA values. To get something useful, 
we must first convert to a factor, and then convert to a numeric to get numeric
categorical data, which is very useful in situations where the data will be used
for data analysis/machine learning.

```{r}
data8
data8n = as.numeric(data8)
data8n
data8c = as.numeric(as.factor(data8))
data8c
```


## Examining Data Structure

```{r}
str(grass2)
str(grass2[1])
```
```{r}
birds <- matrix(c(47,10,40,2,2,19,3,5,0,2,50,0,10,7,0,46,16,8,4,0,9,3,0,0,2,4,0,6,0,0), nrow = 6, ncol = 5)
colnames(birds) = c("Garden","Hedgerow","Parkland","Pasture","Woodland")
rownames(birds) = c("Blackbird","Caffinch","Great Tit","House Sparrow","Robin","Song Thrush")

str(birds)
```
List path objects and show their structure
```{r}
ls.str(pattern = 'path')
```
See what type of data structure the data is:
```{r}
class(grass2)
class(c(grass2[2]))
class(birds)
class(day1)
class(data7i)
```
## Working with History Commands
### Using History Files

#### Viewing the previous command history
```{r}
history(max.show = 25)
```

####Saving and Recalling Lists of Commands
These don't work when building out the HTML for result so they've been commented
out.
```{r}
savehistory(file = 'output/cfichman-history.txt')
loadhistory(file = 'output/cfichman-history.txt')
history()
```

### Editing History Files

You can set the max number of history entries with the following in windows.
```{r}
Sys.setenv('R_HISTSIZE' = 512)
```

## Saving your work in R

I am using projects, so this section won't work for me. But underneath the hood
there's workspace in the project.

#### Saving Named Objects
```{r}
save(grass2, file = 'output/cfichman-save-file-object.Rdata')
save(data7i,data7nt,data7c, file = 'output/cfichman-save-file-list.Rdata')
save(list = ls(pattern = '^data.$'), file = 'output/cfichman-save-file-pattern.Rdata')
```

#### Save ALL THE THINGS
```{r}
save(list = ls(all=TRUE), file = 'output/cfichman-save-all.Rdata')
save.image(file = "output/cfichman-lab2-image.Rdata")
```

### Reading Data Files from Disk

#### Try it out - Save and Read a Binary Data file to and From Disk
```{r}
ls()
savedata = c(9,2,4,6,5,9,2,1,1,7)
save(savedata, file = 'output/cfichman-savedata-test.Rdata')
load(file='output/cfichman-savedata-test.Rdata')
load(file = file.choose())
savedata
```
### Saving Data to Disk as Text Files

```{r}
data7
write(data7, file = 'output/cfichman-lab2-data7.txt', sep = ',')
```
```{r}
write(data7, file = 'output/cfichman-lab2-data7-2.txt', sep = ',', ncolumns = length(data7))
cat(data7, file='output/cfichman-lab2-data7-2.txt')
```

### Writing Matrix and Frame Data to Disk

```{r}
write.table(birds, file='output/cfichman-lab2-birds.txt', row.names = TRUE, sep = ' ', col.names = TRUE)
write.csv(birds, file ='output/cfichman-lab2-birds.csv', row.names = TRUE)
```

### Writing List Objects to Disk

```{r}
dput(data1, file = 'output/cfichman-lab2-object-save.txt')
dget('output/cfichman-lab2-object-save.txt')
```
### Converting List Objects to Data Frames


```{r}
ex_list = list(c(12,15,17,11,15),c(8,9,7,9))
names(ex_list) = c('mow', 'unmow')
ex_list
```


```{r}
str(ex_list)
```

```{r}
ex_stack = stack(ex_list)
names(ex_stack) = c('numbers','sample')
ex_stack
```


```{r}
unstack(ex_stack)
new_list = as.list(unstack(ex_stack))
```
Creating a list from dataframe
```{r}
list2 = as.list(grass2)
names(list2) = c('idx','species','cut')
list2
```



